---
title: "p8106_hw3_sc5826"
author: "Shivalika Chavan"
date: "2026-03-28"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(caret)
library(ISLR)
library(mlbench)
library(tidymodels)
library(rpart) # CART, pruning
library(rpart.plot)
library(party) # conditional inference tree, CIT
library(partykit)
library(pROC)
library(ranger)


knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## 1 
```{r}
college = read_csv("./data/College.csv") |> 
  janitor::clean_names() |> 
  select(-college)

set.seed(1234)
data_split = initial_split(college, prop = 0.8)

training_data = training(data_split)
testing_data = testing(data_split)
n_param = ncol(college) - 1
```

### 1a - Regression Tree
```{r}
tree1 = rpart(formula = outstate ~ . , 
              data = training_data,
              control = rpart.control(cp = 0.008))

rpart.plot(tree1)
```

### 1b - Random Forest
```{r}
ctrl = trainControl(method = "cv", number = 10)

rf.grid = expand.grid(mtry = 1:n_param, # test all possible which is 1 to p (p = 16)
                      splitrule = "variance", # split based on rss, no need to change
                      min.node.size = c(2, 4, 6)) # smallest # of obs in each node
set.seed(1234)
rf.fit = train(outstate ~ . , 
               data = training_data, 
               method = "ranger", # better to use fast implementation
               tuneGrid = rf.grid,
               trControl = ctrl)

ggplot(rf.fit, highlight = TRUE)

set.seed(1234)
rf.final.per = ranger(outstate ~ . , 
                      training_data, 
                      mtry = rf.fit$bestTune[[1]], 
                      min.node.size = rf.fit$bestTune[[3]],
                      splitrule = "variance",
                      importance = "permutation",
                      scale.permutation.importance = TRUE) 

barplot(sort(ranger::importance(rf.final.per), decreasing = FALSE), 
        las = 2, horiz = TRUE, cex.names = 0.7,
        col = colorRampPalette(colors = c("cyan","blue"))(8))

set.seed(1234)
rf.final.imp = ranger(outstate ~ . , 
                      training_data, 
                      mtry = rf.fit$bestTune[[1]], 
                      splitrule = "variance",
                      min.node.size = rf.fit$bestTune[[3]],
                      importance = "impurity") 

barplot(sort(ranger::importance(rf.final.imp), decreasing = FALSE), 
        las = 2, horiz = TRUE, cex.names = 0.7,
        col = colorRampPalette(colors = c("cyan","blue"))(8))

rf_predictions = predict(rf.fit, newdata = testing_data)
test_rmse = RMSE(pred = rf_predictions, obs = testing_data$outstate)
```

The tuned RF model had a test RMSE of `r round(test_rmse, 2)`. Looking at the top 5 variables from both permutation and impurity importance metrics, instructional expenditure per student (`expend`), room and board costs (`room_board`), and percentage of faculty with terminal degrees (`terminal`) all affect tuition costs.  

### 1c - Boosting
```{r}
gbmA.grid = expand.grid(n.trees = c(100,200,500,1000,2000,5000,10000),
                        interaction.depth = 1:5,
                        shrinkage = c(0.001, 0.003, 0.005),
                        n.minobsinnode = 10)
set.seed(1234)
gbmA.fit = train(outstate ~ . , 
                 training_data, 
                 tuneGrid = gbmA.grid,
                 trControl = ctrl,
                 method = "gbm",
                 distribution = "gaussian",
                 metric = "RMSE",
                 verbose = FALSE)

ggplot(gbmA.fit, highlight = TRUE)

summary(gbmA.fit$finalModel, las = 2, cBars = 19, cex.names = 0.6)

gbmA_predictions = predict(gbmA.fit, newdata = testing_data)
test_rmse = RMSE(pred = gbmA_predictions, obs = testing_data$outstate)
```

The tuned model had a test RMSE of `r round(test_rmse, 2)`. Looking at the top 5 variables from this model, instructional expenditure per student (`expend`), room and board costs (`room_board`), and percentage of faculty with terminal degrees (`terminal`) all affect tuition costs. This is in agreement with the random forest models in the previous section. 

## 2 
## Data Cleaning and Test/Train Split:
```{r}
auto_data = read_csv("./data/auto.csv") |> 
  mutate(
    origin = factor(origin, levels = c("1", "2", "3"), 
                    labels = c("American", "European", "Japanese")),
    mpg_cat = factor(mpg_cat, levels = c("low", "high"))
  )

set.seed(1234)
indexTrain = createDataPartition(y = auto_data$mpg_cat, p = 0.7, list = FALSE)
auto_train = auto_data[indexTrain, ] 
auto_test = auto_data[-indexTrain, ] 

n_param = ncol(auto_data) - 1
```

### 2a - Classification Tree
```{r}
set.seed(1234)
tree1 = rpart(formula = mpg_cat ~ . , 
              data = auto_train, 
              control = rpart.control(cp = 0))
rpart.plot(tree1) # biggest possible tree

cptable = tree1$cptable |> as.data.frame()
cptable |> knitr::kable(digits = 4) 
plotcp(tree1)

# minimum cross-validation error; may also use the 1SE rule
minErr = which.min(cptable$xerror)
tree_min = rpart::prune(tree1, cp = cptable$CP[minErr])
rpart.plot(tree_min)

# 1SE Rule
threshold = cptable$xerror[minErr] + cptable$xstd[minErr]
se_row = min(which(cptable$xerror <= threshold))
tree_1se = rpart::prune(tree1, cp = cptable$CP[se_row])
rpart.plot(tree_1se)
```

The tree size corresponding to both lowest cross-validation error and the 1SE rule is 2. This makes sense since there are so few predictors to begin with. 

### 2b - Boosting
```{r}
gbm.grid = expand.grid(n.trees = c(100, 500, 1000, 2000),
                       interaction.depth = 1:5,
                       shrinkage = c(0.001, 0.005, 0.01),
                       n.minobsinnode = 10)

ctrl = trainControl(method = "cv", 
                    classProbs = TRUE, 
                    summaryFunction = twoClassSummary)

set.seed(1234)
model_gbm = train(mpg_cat ~ ., 
                  data = auto_train, 
                  method = "gbm",
                  distribution = "bernoulli",
                  metric = "ROC",
                  tuneGrid = gbm.grid,
                  trControl = ctrl,
                  verbose = FALSE)

summary(model_gbm$finalModel, las = 2, cBars = 10)

gbm_probs = predict(model_gbm, newdata = auto_test, type = "prob")$high
gbm_roc = roc(auto_test$mpg_cat, gbm_probs)
plot(gbm_roc)
auc(gbm_roc)
```

The model achieved an AUC of `r round(auc(gbm_roc), 2)` on the test dataset. This means the model can distinguish well between cars with high and low predicted gas mileage. 